Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-26 paper/ResponseToReviews.tex

We have added text to make this point clearly, and emphasize the associated caveats.
When we first describe the deep mutational scanning approach, we now say:
\begin{quote}
\textsl{
However, in our experiments we mutagenized only the ectodomain and transmembrane domain of Env, and excluded the signal peptide and cytoplasmic tail.
The reason is that we measure how Env mutations affect viral growth, which is influenced both by the functionality of Env protein molecules and their expression level.
Mutations in the signal peptide and cytoplasmic tail commonly affect Env expression level (Chakrabarti et al, 1989; Yuste et al, 2004; Li et al, 1994), so we excluded these regions with the goal of reducing the degree to which we simply identified mutations that affected Env expression.
}
\end{quote}
Then when we describe the results, we clearly emphasize how expression level (like Env protein stability, which was already discussed) could have general effects on the results:
\begin{quote}
\textsl{
Differences in Env's expression level might also contribute to a general broadening or narrowing of tolerance to subsequent mutations.
Because our experiments select for viral growth---which is affected by both Env function and expression---it is possible that some of the shifts are due to epistatic mutational effects on expression rather than function.
}
\end{quote}
}

2. Only the native, pre-fusion Env trimer is used to calculate proximity of substituted and shifted residues. It is possible that substitution-shift colocalization may be detected in other conformations of Env as it undergoes structural rearrangement to mediate fusion, as opposed to acting through longer-range allostery in the pre-fusion conformation. Another well-characterized structure that should be considered is the post-fusion six-helix bundle of gp41.

{\color{black}
This is an excellent suggestion.
In the initial paper, we only analyzed the closed pre-fusion conformation of the Env trimer.
We now analyze substitution-shift co-localization in two other conformations: the open CD4-bound conformation and the post-fusion six-helix bundle.
In the end, our basic conclusions remain the same, adding greater support to the hypothesis that epistasis is acting via long-range interactions.

First, we investigated the open CD4-bound conformation.
Figure 7D statistically tests the hypothesis that shifts are larger at sites that have substituted or sites that contacts substitutions.
Initially, this figure only showed the results of this test for the closed pre-fusion conformation.
Now, it shows the results for both the closed and open conformations.
Both conformations gave the same general result: there is no significant trend for shifted sites to contact sites of substitutions.
We also considered both structures at once, classifying sites as contacting substitutions if they do so in at least one structure.
Even then, we did not find an association (see the new right-most panel in Figure 7D).
Thus, our investigation of the open conformation re-enforced our initial conclusion that many of the shifts are likely due to long-range, rather than short-range, epistatic interactions.

We also repeated the analyses in Figure 7B and Figure 7C with the open conformation.
Our initial conclusions about the solvent accessibility (Figure 7B) and clustering of shifted sites (Figure 7B) also held for the open conformation.

Next, we investigated substitution-shift co-localization in the post-fusion six-helix bundle.
Only $\sim$80 residues per Env monomer are resolved in available structures of the six-helix bundle.
The structure with largest number of shifted sites (PDB: 1ENV) only contains four shifted sites and five substituted sites.
These numbers are too low for the statistical tests used in Figure 7.
But we have added a new figure (Figure 7 - Figure supplement 2) that qualitatively analyzes substitution-shift co-localization in the post-fusion six-helix bundle.
Three of the four shifted sites cluster at one end of the helix, along with one of the five substitutions.
However, this cluster is also present in the closed and open states as well (Figure 7 - Figure supplement 2).
Thus, the post-fusion six-helix bundle did not reveal any new patterns of co-localization that were not also present in the other states.

We now describe the results of the analyses in the first two paragraphs of the Results subsection ``Structural and evolutionary properties of shifted sites", interleaved with the original text in these paragraphs.

In the process of making these revisions, we discovered a minor bug that affected our calculations of residue-residue distances. 
We fixed that bug in the revised version. 
None of the main results change, but some of the quantitative values in the box plots and the corresponding P-values have changed slightly.
}

@@

3. Overall, it is found that the shifted sites tend to cluster together, but they are not necessarily the same as the sites with substitutions between the two Envs or ones that contact them. This analysis is not pursued in great detail, however, leaving the reasons why shifted sites tend to cluster together unexplained. Given that the authors also observe some evidence of entrenchment of substitutions between the Envs, is it possible that interactions between the shifted sites themselves are important? This question could be explored through a more detailed examination of the structure.

{\color{black}
To address this comment, we examined the clusters of shifted sites in greater detail.
Two clusters occur within conformationally dynamic regions of Env.
One hypothesis as to why shifted sites cluster in these regions is that their conformationally dynamic nature allows long-range epistatic interactions to be propagated between shifted and substituted sites that are distant from one another.

We have added a new supplemental figure (Figure 7 - figure supplement 3) that provides a fine-grained structural view of each of the clusters.
We have also added an entire paragraph detailing these observations at the end of the section in the Results called ``Structural and evolutionary properties of shifted sites".
Overall, these additions strengthen the manuscript by providing a better structural intuition for how long-range epistatic interactions might occur in Env.
}

@@

3. Page 6: The dN/dS parameter is introduced, but its meaning in the two different models is only described in the next paragraph. Noting that a description follows may help readers.

{\color{black}
This is a good suggestion.
We have re-structured the relevant text (second and third paragraph of third subsection of Results) to expand the explanation of the dN/dS model parameters and put them in a more logical order.}

4. The color scheme of structures in Fig 7a, 9a is difficult to discern. Higher contrast colors please.

{\color{black}
We have now changed the color schemes to be higher contrast in both of these figures.
The one in Fig 7a now goes from white-to-red instead of grey-to-red.
And the one in Fig 9a now goes from blue-to-white-to-red instead of blue-to-grey-to-red.}

### 2018-02-26 paper/manuscript.tex
fixed typo on Adam commit (`may` was repeated twice in a row)`

Indeed, many of the shifted sites cluster within regions of Env that are highly conformationally dynamic.
\FIGSUPP[shiftfeatures]{dynamic_clusters} shows the structural context of these clusters in finer detail.
One cluster is at the trimer apex where two of Env's variable loops pack against one another and against an adjacent protomer.
These interactions are likely involved in regulating the transition between conformational states, and upon CD4 binding, these loops become highly disordered~\citep{guttman2014cd4,ozorowski2017open}.
Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\citep{lee2017broadly}.
Strikingly, this cluster of shifted sites may may reflect previously observed differences in the conformational dynamics of this regions between these two Envs; the V2 region of BF520 SOSIP trimer is more accessible to deuterium exchange than the BG505 SOSIP trimer~\citep{verkerke2016epitope}.
The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\citep{ozorowski2017open}.
One of the shifted sites (site 69) overlaps with this network, and mutations at another (site 64) have been shown to strongly modulate the relative stability of the open and closed conformations~\citep{de2015immunogenicity}.
In total, these two clusters consist of nearly half of the shifted sites (13 out of 30).
One hypothesis why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
It is difficult to discern exactly how these interactions might occur, but there is certainly a trend for sites that are conformationally dynamic to also be sites that show shifts in their amino-acid preferences during evolution.

>>>>

Indeed, many of the shifted sites cluster within regions of Env that are highly conformationally dynamic.
\FIGSUPP[shiftfeatures]{dynamic_clusters} shows the structural context of these clusters in finer detail.
One cluster is at the trimer apex where two of Env's variable loops pack against one another and against an adjacent protomer.
These interactions are likely involved in regulating the transition between conformational states, and upon CD4 binding, these loops become highly disordered~\citep{guttman2014cd4,ozorowski2017open}.
Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\citep{lee2017broadly}.
Strikingly, this cluster of shifted sites may reflect previously observed differences in the conformational dynamics of this regions between these two Envs; the V2 region of BF520 SOSIP trimer is more accessible to deuterium exchange than the BG505 SOSIP trimer~\citep{verkerke2016epitope}.
The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\citep{ozorowski2017open}.
One of the shifted sites (site 69) overlaps with this network, and mutations at another (site 64) have been shown to strongly modulate the relative stability of the open and closed conformations~\citep{de2015immunogenicity}.
In total, these two clusters consist of nearly half of the shifted sites (13 out of 30).
One hypothesis why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
It is difficult to discern exactly how these interactions might occur, but there is certainly a trend for sites that are conformationally dynamic to also be sites that show shifts in their amino-acid preferences during evolution.
